### 2010-02-24 sarahDraft.tex
Types of Occupancy Models includes MLE

		Occupancy is "The probability that a randomly selected site or sampling 
unit in an area of interest is occupied by a species" [MacKenzie2006] This is 
not the same as the abundance, or population size, of a species.  Instead of 
measuring the number of animals present in a species, occupancy measures the 
proportion of land that is used as the habitat of at least one member of the 
species. 
	is either the probability or the proportion 25-  note importance of the 
distinction although they can be considered the same thing in most cases

		*What is it used for?

	\subsection{Different Ways of Sampling Data}
		Most common way of sampling:  visit sites and spend time within each one 
looking for either individuals of the species or for evidence the species is 
present- "presence-absence survey"(26).  can use visual observation of animals,
capture animals, ovserve tracks, detect vocalizations, camera traps, sound 
recording.  leads to naive occupancy rate.  PROBLEM: false negatives.

		*Measuring rate of occupancy vs. population \\
		*Looking at tracks vs. listening to stuff or acutally looking for the 
			animal

	\subsection{Assumptions}
		*Independence of detection between sites \\
		*No false positives \\
		*Presence and detection probabilities constant across sites and surveys \\
		*Closed population

	\subsection{Types of Occupancy Models}
		We only considered single-season, single-species models because this is 
the data we had, but any of these models can be extended to take into account
multiple seasons or multiple species.  
		Naive Occupancy pg 11- this is when you just divide the number of sites
			that are indicated as occupied by the total number of sites.  it is
			generally accepted that this is too low because not all animals are 
			detected.
		Hierarchical Models:
		Maximum Likelihood \\  Give the simple model, then say it can be 
			modified in many different ways.  give some examples possibly 
			without going into detail.
		Any other models? \\
		Bayesian?  is that the name of what we used?  should save most of this 
			for theoretical section?

>>>>

		Occupancy has two similar definitions.  It can be "the probability that a
randomly selected site or sampling unit in an area of interest is occupied by a 
species" [MacKenzie2006].  It also can be used to describe "the proportion of 
area, patches, or sample units that is occupied" [MacKenzie2006].  Occupancy is 
not the same as the abundance, or population size, of a species.  The number of
members of a species is not important, only whether or not at least one member of
the species is present in each site.

	While there is a difference between these two definitions of occupancy, in 
general the probability is unknown and the observed proportion is used as an 
estimate of the underlying probability.  Because of this, these two definitions
of occupancy are oftentimes used interchangably.

	The main reasons that people are interested in measuring occupancy are science
or conservation and management [MacKenzie2006].  This statistic can give insight 
into the way a system works.  It also can be used to monitor animals habitats 
and discover trends over time in animal populations.

	\subsection{Different Ways of Sampling Data}

		In order to determine occupancy rate, the data is often collected in the
 form of a presence-absence survey.  This means that observers visit sites and 
spend time within each one looking for the animals themselves, tracks, 
vocalizations, or some other indicator that the animal is present.  In some cases, 
camera traps and sound recording are used instead of human observers.

	The main problem with these methods of sampling data is that it is possible 
for an animal to occupy a site, but not be detected by the observer.  This means
that the site is recorded as absent, but there really is an animal there.  This 
is called a false negative.  Marking an animal present when there really is no
animal is also a possible problem, but is thought to be much less likely and is
often ignored.  We will discuss this more below.

	\subsection{Assumptions}

	There are some assumptions that are made in order to create models of 
occupancy.  Without these assumptions, the models become much more complicated.
One common assumption is no false positives, which means if a site is recorded as
present, this cannot be incorrect.  This is an assumption that we attempted to 
get rid of in our models.  A second common assumption is that presence and 
detection probabilties are constant across sites and surveys.  A third common 
assumption is that there is independence of detection between sites.  A fourth
common assumption is that the surveys are being conducted on a closed population.
This means that no animals leave the area or join the area during the time of the
surveys.

	Obviously, not all of these assumptions are always true.  In some cases, there
are established ways to deal with getting rid of an assumption in the models, 
and in others there are no common ways to get rid of the assumption.  Either way, 
it adds a level of complication to the model, and therefore we will only talk 
about dealing with the assumption of no false positives, and independence of 
detection across sites.

	\subsection{Types of Occupancy Models}

		All of the models that we looked at were single-species, single-season
models because this is what fit our data, but any of these models can be modified
to take into account multiple seasons or multiple species.

	The most simple way to measure occupancy is to simply divide the number 
of sites where an animal was found to be present by the total number of sites.  
This is called naive occupancy [MacKenzie2006].  It is generally accepted that 
this estimate is too low because animals are not detected at all sites where they
are present, which is the issue of false negatives.  The other occupancy models
attempt to do better than this simple estimate.

	One way to create more accurate models is to use a Hierarchical model to 
think about the parameters that affect occupancy.

	A more complicated model that can be used as a Hierarchical model is Maximum
Likelihood Estimation (MLE).  This is done by finding the probability 
distribution of the observed data as a function of the parameters.  This 
distribution is then said to be the function of the parameters conditional on the
data, and becomes the likelihood function.  Then, the values of the parameters 
that maximize the likelihood function are found, which means that we find the 
parameters that make the data most likely given the underlying model.  
[MacKenzie2006]

	Another way to do this is using a Bayesian model, which we will discuss in 
more detail below.
